Statistical quantification of confounding bias in predictive modelling

### 2021-10-30 manuscript.tex

% keywords can be removed
\keywords{machine learning, predictive modelling, confoudner bias, confounder test, conditional independence, conditional permutation}

@@

Spurious, out-of-interest associations between the predictor variables (features) and the prediction target can be detrimental to the model's biomedical validity and generalizability. This phenomenon is often called confounding bias\citep{prosperi2020causal}. Confounding bias can be driven by - among others - measurement artifacts (e.g motion artifacts in magnetic resonance imaging-based predictive models of Alzheimer's\citep{rao2017predictive}, attention deficit hyperactivity disorder\citep{eloyan2012automated, couvy2016head} or Autism Spectrum Disorder (ASD)\citep{gotts2013perils, spisak2014voxel, spisak2019optimal}), demographic and psychometric variables (e.g models trained to predict intelligence\citep{ cole2012global, he2020deep} might provide a statistically significant predictive performance by picking up solely on age-related variance\citep{dubois2018distributed, lohmann2021predicting}), sampling bias and stochastic group differences (e.g. racially biased machine learning models\citep{ obermeyer2019dissecting, lwowski2021risk}), batch effects or, in multi-center studies, center-effects.

@@

Powerful and robust statistical tests for quantifying confounding bias in predictive models could substantially foster both the identification of confounders to correct for and the assessment of the effectiveness of various confound-mitigation approaches. Although it is tempting to think about confounding bias as the \emph{conditional dependence} of the model predictions on the confounder, given the target variable, the proper evaluation of the conditional independence among these variables is challenging. Namely, even in the presence of a slight non-normality and/or non-linearity of the involved conditional distributions, the partial correlation analogs of bivariate non-parametric correlations (like the partial Spearman correlation, see Fig. \ref{fig:sim-h0-demo}) are not valid measures of conditional independence. Although warnings about this issue were given from early on\citep{korn1984ranges}, and received a fair amount of attention recently\citep{bergsma2010nonparametric, candes2016panning, peters2016causal,  shah2020hardness, berrett2020conditional}, the magnitude of the problem may not be fully appreciated in case of predictive model diagnostics, where non-normality and non-linearity of the model output can be frequently seen, as a consequence of e.g. feature-set characteristics and model regularization\citep{garcia2009study, kristensen2017whole} (see Supplementary Material \ref{sup:nomlinviol} for a simplistic example).

@@

This work aims to construct a statistical test for confounding bias that (i) can be applied post-hoc, without re-fitting the model (i.e. requires only the target variable, the confounder and the model predictions), (ii) guarantees valid type-I error control even if the involved conditional distributions are non-normal and/or non-linear, (iii) is applicable for classification as well as for prediction problems, both with numerical and categorical confounders.

%fig:sim-h0-demo
\begin{figure}[!b]
  \centering
  \includegraphics[width=0.5\paperwidth]{fig/sim_h0_demo.eps}
  \caption{\textbf{Type I error control of partial Spearman correlation, linear and GAM-based conditional permutation test}. \\
  Type I error control was investigated in three example cases: normal conditional distribution with linear dependency (first row), slightly non-normal conditional distribution with linear dependency (second row) and normal conditional distribution with non-normal (sigmoid) dependency (third row). Non-normal conditional distribution on the second plot is illustrated with blue density diagrams (kurtosis: -0.8, skewness: -0.1). False positive rates for various simulation settings is shown in heatmaps. The upper limit for the binomial confidence interval corresponding to $alpha=0.05$ is 0.065. Values below this threshold indicate a valid type I error control and colored white.
  }
  \label{fig:sim-h0-demo}
\end{figure}

@@

% fig: overview
\begin{figure}[!b]
  \centering
  \includegraphics[width=0.6\paperwidth]{fig/overview.eps}
  \caption{\textbf{Overview of the proposed partial confounder test}. \\
    The partial confounder test models the conditional distribution of the confounder, given the target variable, with a generalized additive model (GAM). The parallel-pairwise Markov-chain Monte-Carlo sampler draws permutations of the confounder variable that comply with the GAM-based conditional distribution (permutation 1, 2, ..., m). The test statistic (coefficient of determination, $R^2$) is then computed between the model predictions and the original, as well as the permuted confounder variables. The original and the permuted test statistics construct the p-value as the ratio of permuted test statistics more extreme than the original. Figure source code available as jupyter notebook: \href{https://github.com/pni-lab/mlconfound-manuscript/blob/main/simulated/overview-fig.ipynb}{https://github.com/pni-lab/mlconfound-manuscript/blob/main/simulated/overview-fig.ipynb}
  }
  \label{fig:overview}
\end{figure}

@@

The idea of conditional independence provides a straightforward framework for assessing confounding bias in predictive models. However, handling the non-normal and/or non-linear conditional dependence often seen in predictive models\citep{garcia2009study, kristensen2017whole} poses a great challenge.
In fact, as recently shown by Shah and colleagues\cite{shah2020hardness} in their 'no free lunch' theorem, it is effectively impossible to establish a fully non-parametric conditional independence test with a valid type I error control and a non-trivial power. In practice, this means that no fully assumption-free test for assessing confounding bias can be constructed. Indeed - perhaps somewhat surprisingly, but not totally unexpectedly\cite{korn1984ranges} - partial correlation-like analogs of the widely used bivariate non-parametric test, like partial Spearman correlation, exhibits inflated type I errors even with slight violations of normality and/or linearity (as clearly demonstrated with simulated data on Fig. \ref{fig:sim-h0-demo}). Such tests are therefore poor choices for testing confounding bias in machine learning.

As, in terms of its conditional distribution on the others, the most intractable from the three involved variables (target, prediction, confounder) is clearly the model output\citep{garcia2009study, kristensen2017whole}, a method being distribution-free only for this variable might already provide a sufficient robustness for predictive model diagnostics. Exactly this can be achieved with the proposed approach, which extends the novel framework of conditional permutation testing (CPT)\cite{berrett2020conditional} with conditional distribution estimation via generalized additive (GAM)\citep{hastie1987generalized} or multinomial logistic models\cite{bennett1966multiple, jones1975proability}. The proposed approach offers two novel tests for probing confounding bias: the \emph{full confounder test} probes whether the model's predictive performance can be exclusively attributed to the confounder and the \emph{partial confounder test} investigates whether the model utilizes any confounder-variance in the predictions, when controlled for the target variable.
These tests place no assumptions on the conditional distributions of the model predictions, ensuring valid model diagnostics even in cases of non-normally and non-linearly dependent predictions.

The inner workings of the \emph{partial confounder test} is summarized on Fig. \ref{fig:overview}. In short, the test models the conditional distribution between the confounder and the target variable with a generalized additive model (GAM) - or with a multinomial logistic regression, in case of categorical confounder - and then uses a so-called parallel-pairwise Markov-chain Monte-Carlo sampler that draws permutations of the confounder, so that the permuted variables still comply with the GAM-based conditional distribution. The test statistic (coefficient of determination, $R^2$) is then computed between the model predictions and the original, as well as the permuted variables. The original and the permuted test statistics construct the p-value as the ratio of permuted test statistics more extreme than the original. The \emph{full confounder test} works in an analogous way, with the difference that it creates permuted copies of the target variable, instead of the confounder.

@@

%fig:sim-normal
\begin{figure}[!b]
  \centering
  \includegraphics[width=0.6\paperwidth]{fig/sim_normal.eps}
  \caption{\textbf{Type I error control and power of the partial confounder test based on simulations with normal conditional distribution and linear dependencies.} \\
  Heatmaps depict positive rates (ratio of p-values lower than 0.05, color coded as shown by the palette on the right) in various simulations settings (100 simulations per tile) with different simulation weights $w_{y\hat{y}}$ (horizontal axis on each heatmap), $w_{yc}$ (vertical axis on each heatmap), $w_{c\hat{y}}$ (rows) and for different sample sizes (N, columns). Weights 0.2, 0.33, 0.4, 0.6, 0.66, 1.0 can be assigned to the following approximate explained variance values: 4\%, 10\%, 12\%, 25\%, 30\%, 50\%. First row contains simulations under the null hypothesis (H0, no confounder effect), rows 2-4 represent simulations from the alternative hypothesis (H1, confounding bias).
  Positive rates for the simulations under the null and the alternative hypotheses can be interpreted as type I error rate and statistical power, respectively. The higher 95\% confidence limit for a positive rate of $alpha=0.05$ is 0.11 for each tile.
  }
  \label{fig:sim-normal}
\end{figure}

@@

Simulations show that non-normality has minimal effect on the power of the tests, except in case of extreme non-normality. (Fig. \ref{fig:sim-non-normal}). Simulations with sigmoid dependence resulted in an apparent loss of statistical power, however this is simply a consequence of the simulation methodology: with the same parameters, the sigmoid transformed confounder explains only approximately half the variance as compared to linear simulations.
Type I error control was valid in case of categorical variables and for the \emph{full} confounder test, as well. Power in these cases was also highly similar to the partial confoudner test results with numerical variables. (Supplementary Figures \ref{fig:sim-bbb-lin-partial}, \ref{fig:sim-bbb-lin-full}, \ref{fig:sim-bbb-sig-partial} and \ref{fig:sim-bbb-sig-full}).

@@

\end{tabular}
\caption{\label{tab:unconditional-pvals} Coefficients-of-determination ($R^2$), the corresponding p-values and the p-values of the partial and full confounder tests, for all investigated confounders (conf.) confounder-mitigation methods (method). Bold numbers denote significant confounding bias identified by the partial confounder test or the lack of significant contribution over that of the confoudner, as identified by the full confounder test.}
\end{table}

@@

In the following section, CPT is adapted for partial confounder testing (option 3) and extended with general additive model\cite{hastie1987generalized} (GAM) and multinomial logistic regression\citep{bennett1966multiple, jones1975proability} based conditional distribution estimations, in order to make it handle categorical data and non-linear dependencies between the confoudner and the target variable. (For an overview of the method, see Fig. \ref{fig:overview}). The formulation of the full confounder test (option 1) is analogous and given in Supplementary Material \ref{sup:full-test}.

@@

Note that simulations with $w_{c\hat{y}}=0$ produce data under the null hypothesis of no confoudner bias.

>>>>

% keywords can be removed
\keywords{machine learning, predictive modelling, confounding bias, confounder test, conditional independence, conditional permutation}

@@

Spurious, out-of-interest associations between the predictor variables (features) and the prediction target can be detrimental to the model's biomedical validity and generalizability. This phenomenon is often called confounding bias\citep{prosperi2020causal}. Confounding bias can be driven by - among others - measurement artifacts (e.g motion artifacts in magnetic resonance imaging-based predictive models of Alzheimer's\citep{rao2017predictive}, attention deficit hyperactivity disorder\citep{eloyan2012automated, couvy2016head} or Autism Spectrum Disorder (ASD)\citep{gotts2013perils, spisak2014voxel, spisak2019optimal}), demographic and psychometric variables (e.g models trained to predict intelligence\citep{ cole2012global, he2020deep} might provide a statistically significant predictive performance by picking up solely on age-related variance\citep{dubois2018distributed, lohmann2021predicting}), sampling bias and stochastic group differences (e.g. racially biased machine learning models\citep{ obermeyer2019dissecting, lwowski2021risk}) as well as batch effects or, in multi-center studies, center-effects.

@@

Powerful and robust statistical tests for quantifying confounding bias in predictive models could substantially foster both the identification of confounders to correct for and the assessment of the effectiveness of various confound-mitigation approaches. It is tempting to think about confounding bias as the \emph{conditional dependence} of the model predictions on the confounder, given the target variable. However, the proper evaluation of conditional independence among these variables is challenging. Namely, even in the presence of a slight non-normality and/or non-linearity of the involved conditional distributions, the 'conditional' analogs of the most popular bivariate non-parametric tests (like the partial Spearman correlation, see Fig. \ref{fig:sim-h0-demo}) are not valid measures of conditional independence. Although warnings about this issue were given from early on\citep{korn1984ranges}, and received a fair amount of attention recently\citep{bergsma2010nonparametric, candes2016panning, peters2016causal,  shah2020hardness, berrett2020conditional}, the magnitude of the problem may not be fully appreciated in case of predictive model diagnostics, where non-normality and non-linearity of the model output can be frequently seen, as a consequence of e.g. feature-set characteristics and model regularization\citep{garcia2009study, kristensen2017whole} (see Supplementary Material \ref{sup:nomlinviol} for a simplistic example).

@@

This work aims to construct a statistical test for confounding bias that (i) guarantees valid type-I error control for arbitrary models, even if non-normal and/or non-linear dependencies are involved (ii) does not require re-fitting the model, (iii) is applicable for classification as well as for prediction problems and both with numerical and categorical confounders.

%fig:sim-h0-demo
\begin{figure}[!b]
  \centering
  \includegraphics[width=0.5\paperwidth]{fig/sim_h0_demo.eps}
  \caption{\textbf{Type I error control of partial Spearman correlation, linear and GAM-based conditional permutation test}. \\
  Type I error control was investigated in three example cases: normal conditional distribution with linear dependency (first row), slightly non-normal conditional distribution with linear dependency (second row) and normal conditional distribution with non-normal (sigmoid) dependency (third row). Non-normal conditional distribution on the second plot is illustrated with blue density diagrams (kurtosis: -0.8, skewness: -0.1). False positive rates for confounder contributions ($w_{yc}$) and predictive performances ($w_{y\hat{y}}$) is shown in heatmaps. The upper limit for the binomial confidence interval corresponding to $alpha=0.05$ is 0.065. Values below this threshold (colored white) indicate a valid type I error control.
  }
  \label{fig:sim-h0-demo}
\end{figure}

@@

% fig: overview
\begin{figure}[!b]
  \centering
  \includegraphics[width=0.6\paperwidth]{fig/overview.eps}
  \caption{\textbf{Graphical representation of the proposed partial confounder test}. \\
    The partial confounder test models the conditional distribution of the confounder, given the target variable, with a generalized additive model (GAM). The parallel-pairwise Markov-chain Monte-Carlo (MCMC) sampler draws permutations of the confounder variable that comply with the GAM-based conditional distribution (permutation 1, 2, ..., m). The test statistic (coefficient of determination, $R^2$) is then computed between the model predictions and the original, as well as the permuted confounder variables. The original and the permuted test statistics construct the p-value as the ratio of permuted test statistics more extreme than the original. Figure source code available as jupyter notebook: \href{https://github.com/pni-lab/mlconfound-manuscript/blob/main/simulated/overview-fig.ipynb}{https://github.com/pni-lab/mlconfound-manuscript/blob/main/simulated/overview-fig.ipynb}
  }
  \label{fig:overview}
\end{figure}

@@

The concept of conditional independence provides a straightforward framework for assessing confounding bias in predictive models. However, handling the non-normal and/or non-linear conditional dependencies often seen in predictive models\citep{garcia2009study, kristensen2017whole} poses a great challenge.
In fact, as recently shown by Shah and colleagues in their 'no free lunch' theorem\cite{shah2020hardness}, it is effectively impossible to establish a \emph{fully non-parametric} conditional independence test with a valid type I error control and a non-trivial power. Indeed, perhaps somewhat surprisingly, but not totally unexpectedly\cite{korn1984ranges} - partial correlation-like analogs of widely used bivariate non-parametric test, like partial Spearman correlation, exhibit inflated type I errors even with slight violations of normality and/or linearity (as clearly demonstrated with simulated data on Fig. \ref{fig:sim-h0-demo}). Such tests are therefore poor choices for testing confounding bias in machine learning.

As, in terms of its conditional distribution on the others, the model output is clearly the most intractable from the three involved variables (target, prediction, confounder)\citep{garcia2009study, kristensen2017whole}, a method being distribution-free only for this variable might already provide a sufficient robustness for predictive model diagnostics. Exactly this can be achieved with the proposed approach, which extends the novel framework of conditional permutation testing (CPT)\cite{berrett2020conditional} with conditional distribution estimation via generalized additive (GAM)\citep{hastie1987generalized} or multinomial logistic models (mnlogit)\cite{bennett1966multiple, jones1975proability}. The proposed approach offers two novel tests for probing confounding bias: the \emph{full confounder test} probes whether the model's predictive performance can be exclusively attributed to the confounder and the \emph{partial confounder test} investigates whether the model utilizes any confounder-variance in the predictions, when controlled for the target variable.
These tests place no assumptions on the conditional distributions of the model predictions, ensuring valid model diagnostics even in cases of non-normally and non-linearly dependent predictions.

The inner workings of the \emph{partial confounder test} are summarized on Fig. \ref{fig:overview}. In short, the test models the conditional distribution between the confounder and the target variable with a GAM - or with an \emph{mnlogit} regression, in case of categorical confounder - and then uses a so-called parallel-pairwise Markov-chain Monte-Carlo sampler\cite{berrett2020conditional} that draws permutations of the confounder, so that the permuted variables still comply with the estimated conditional distribution. The test statistic (coefficient of determination, $R^2$) is then computed between the model predictions and the original, as well as the permuted variables. The original and the permuted test statistics construct the p-value as the ratio of permuted test statistics more extreme than the original. The \emph{full confounder test} works in an analogous way, with the difference that it creates permuted copies of the target variable, instead of the confounder.

@@

%fig:sim-normal
\begin{figure}[!b]
  \centering
  \includegraphics[width=0.6\paperwidth]{fig/sim_normal.eps}
  \caption{\textbf{Type I error control and power of the partial confounder test based on simulations with normal conditional distribution and linear dependencies.} \\
  Heatmaps depict positive rates (ratio of p-values lower than 0.05, color coded as shown by the palette on the right) in various simulations settings (100 simulations per tile) with different simulation weights $w_{y\hat{y}}$ (predictive performance; horizontal axis on each heatmap), $w_{yc}$ (confounder-target association; vertical axis on each heatmap), $w_{c\hat{y}}$ (degree of confoudner bias; rows) and for different sample sizes (N, columns). Weights 0.2, 0.33, 0.4, 0.6, 0.66, 1.0 can be assigned to the following approximate explained variance values: 4\%, 10\%, 12\%, 25\%, 30\%, 50\%, respectively. First row contains simulations under the null hypothesis (H0, no confounder bias), rows 2-4 represent simulations from the alternative hypothesis (H1, confounding bias).
  Positive rates for the simulations under the null and the alternative hypotheses can be interpreted as type I error rate and statistical power, respectively. The higher 95\% confidence limit for a positive rate of $alpha=0.05$ is 0.11 for each tile.
  }
  \label{fig:sim-normal}
\end{figure}

@@

Simulations show that non-normality has minimal effect on the power of the tests, except in case of extreme non-normality. (Fig. \ref{fig:sim-non-normal}). Simulations with sigmoid dependence resulted in an apparent loss of statistical power, however this is simply a consequence of the simulation methodology: with the same parameters, the sigmoid transformed confounder explains only approximately half the variance as compared to linear simulations.
Type I error control was valid in case of categorical variables and for the \emph{full} confounder test, as well. Power in these cases was also highly similar to the partial confounder test results with numerical variables. (Supplementary Figures \ref{fig:sim-bbb-lin-partial}, \ref{fig:sim-bbb-lin-full}, \ref{fig:sim-bbb-sig-partial} and \ref{fig:sim-bbb-sig-full}).

@@

\end{tabular}
\caption{\label{tab:unconditional-pvals} Coefficients-of-determination ($R^2$), the corresponding p-values and the p-values of the partial and full confounder tests, for all investigated confounders (conf.) confounder-mitigation methods (method). Bold numbers denote significant confounding bias identified by the partial confounder test or the lack of significant contribution over that of the confounder, as identified by the full confounder test.}
\end{table}

@@

In the following section, CPT is adapted for partial confounder testing (option 3) and extended with general additive model\cite{hastie1987generalized} (GAM) and multinomial logistic regression\citep{bennett1966multiple, jones1975proability} based conditional distribution estimations, in order to make it handle categorical data and non-linear dependencies between the confounder and the target variable. (For an overview of the method, see Fig. \ref{fig:overview}). The formulation of the full confounder test (option 1) is analogous and given in Supplementary Material \ref{sup:full-test}.

@@

Note that simulations with $w_{c\hat{y}}=0$ produce data under the null hypothesis of no confounder bias.
